---
format: pdf
authors: Suchanut Namcharoen, Ashish Job, Chloe Li, Emma Pham, Jaden Thai 
editor: visual
---

\begin{center}
      \vspace{5pt}
      \textbf{The Report on Using Machine Learning Models to Find the Quality Of Sleep}\\ \vspace{5pt}
      Group 5: Suchanut Namcharoen, Ashish Job, Chloe Li, Emma Pham, Jaden Thai \\ \vspace{5pt}
      University of Houston \\ \vspace{5pt}
      MATH 4322 - Machine Learning \\ \vspace{5pt}
      Dr. Poliak \\ \vspace{5pt}
      April 27, 2024
\end{center}

\pagebreak

# Introduction

## Background

\quad\quad Sleep is undeniably one of the most critical aspects of our health, particularly for college students. Balancing academic responsibilities, social life, and possibly part-time jobs or internships often leaves us feeling overwhelmed. Despite our best efforts to manage our time efficiently, adequate sleep always seems to elude us. Exploring the various factors influencing our sleep quality, including but not limited to our commitments, could provide valuable insights into this ongoing struggle. As a result, we decided to look into just that, and we came across the \textit{Sleep Health and Lifestyle Dataset} on \textit{Kaggle} that "comprises 400 rows and 13 columns, covering a wide range of variables related to sleep and daily habits."

## Our Dataset

\quad\quad As previously mentioned, the **Sleep Health and Lifestyle Data set** contains 400 rows and 13 columns (400 observations and 13 variables) of data that relate to an individual's lifestyle and sleep habits. The variables are as follows:

-   Person ID: An identifier for each individual.
-   Gender: The gender of the person (Male/Female).
-   Age: The age of the person in years.
-   Occupation: The occupation or profession of the person.
-   Sleep Duration (hours): The number of hours the person sleeps per day.
-   Quality of Sleep (scale: 1-10): A subjective rating of the quality of sleep, ranging from 1 to 10.
-   Physical Activity Level (minutes/day): The number of minutes the person engages in physical activity daily.
-   Stress Level (scale: 1-10): A subjective rating of the stress level experienced by the person, ranging from 1 to 10.
-   BMI Category: The BMI category of the person (e.g., Underweight, Normal, Overweight).
-   Blood Pressure (systolic/diastolic): The blood pressure measurement of the person, indicated as systolic pressure over diastolic pressure.
-   Heart Rate (bpm): The resting heart rate of the person in beats per minute.
-   Daily Steps: The number of steps the person takes per day.
-   Sleep Disorder: The presence or absence of a sleep disorder in the person (None, Insomnia, Sleep Apnea).
    -   None: The individual does not exhibit any specific sleep disorder.

    -   Insomnia: The individual experiences difficulty falling asleep or staying asleep, leading to inadequate or poor-quality sleep.

    -   Sleep Apnea: The individual suffers from pauses in breathing during sleep, resulting in disrupted sleep patterns and potential health risks.
    
## Our Question    
    
\quad\quad From these variables/columns, we decided to work towards answering the following question: \textbf{Can we predict the quality of sleep (1-10) based on the lifestyle factors present in the dataset?}

## Data Cleaning

\quad\quad To move forward with creating our models to predict the `Quality of Sleep` variable, we realized that we may have to format, or clean, our dataset up, and to clean it up, we evaluated the dataset and determined which variables could possibly interfere with the accuracy of the model. We decided to remove the `PersonID` variable when creating the model, since that represents each observation and would not be used as neither a predictor nor response as well as the `Blood Pressure` variable because we were unable to accurately interpret the blood pressure measurements.

```{r}
# Importing the dataset and viewing summary
data <- read.csv("Sleep_health_and_lifestyle_dataset.csv")
summary(data)
```

\quad\quad We then converted the remaining categorical variables, which included `Occupation`, `Gender`, `Sleep Disorder`, and `BMI Category`.

```{r}
data$Occupation <- as.factor(data$Occupation) 
data$Gender <- as.factor(data$Gender)
data$Sleep.Disorder <- as.factor(data$Sleep.Disorder)
data$BMI.Category <- as.factor(data$BMI.Category)
```

## Our Chosen Approaches

\quad\quad With our data cleaned up, we now had to transition into choosing what models we would be using to answer our proposed question. First, since we were predicting a quantitative value, we already had our options whittled down to a fewer number of options, of which we decided to choose the \textbf{Linear Regression Model}, the \textbf{Decision Tree Model}, and the \textbf{Random Forest Model}.

# Our Methods

## Linear Regression Model

### Why We Chose this Model

\quad\quad We decided to first approach the data set using a multiple linear regression model because we wanted to inference the quality of sleep (1-10), a quantitative response variable, using multiple predictors. Furthermore, linear regression models are generally more straightforward to understand and interpret, so we can quickly identify which predictors are statistically significant and exactly how they affect the response variable.

### Our Formula

\quad\quad In the given equation, $y$ represents the response variable Quality of Sleep, $\beta_0$ represents the model intercept, $\beta_i$ for $i = 1, \ldots, 11$ represents the coefficients of the predictors, where:

\begin{itemize}
    \item \( \beta_1 \) represents the coefficient for Gender,
    \item \( \beta_2 \) represents the coefficient for Age,
    \item \( \beta_3 \) represents the coefficient for Occupation,
    \item \( \beta_4 \) represents the coefficient for Sleep Duration,
    \item \( \beta_5 \) represents the coefficient for Physical Activity Level,
    \item \( \beta_6 \) represents the coefficient for Stress Level,
    \item \( \beta_7 \) represents the coefficient for BMI Category,
    \item \( \beta_8 \) represents the coefficient for Heart Rate,
    \item \( \beta_9 \) represents the coefficient for Daily Steps,
    \item \( \beta_{10} \) represents the coefficient for Daily Steps, and
    \item \( \beta_{11} \) represents the coefficient for Sleep Disorder.
\end{itemize}

So, the equation representing the multiple linear regression model is as follows:

$y$ = $\beta_0$ + $\beta_1$(\text{Gender}) + $\beta_2$(\text{Age}) + $\beta_3$(\text{Occupation}) + $\beta_4$(\text{Sleep Duration}) + $\beta_5$(\text{Physical Activity Level}) + $\beta_6$(\text{Stress Level}) + $\beta_7$(\text{BMI Category}) + $\beta_8$(\text{Heart Rate}) + $\beta_9$(\text{Daily Steps}) + $\beta_{10}$(\text{Daily Steps}) + $\beta_{11}$(\text{Sleep Disorder})

```{r}
# Creating the multiple linear regression model
sleep.model <- lm(Quality.of.Sleep ~.-Blood.Pressure-Person.ID, data = data)
summary(sleep.model)
```

\quad\quad Based on the summary of the linear regression model, we can observe that multiple predictors are statistically significant (low p-values) as indicated by the asterisks on the right side of the predictors. From these predictors, we notice that:

-   Gender Male has a positive estimate (increasing the quality of sleep by approximately 0.56), indicating that males generally have a better quality of sleep compared to females.

-   Each additional year of age is associated with a a 0.057 increase in quality of sleep, suggesting that sleep quality might improve with age.

-   All of the occupations have a negative estimate, with sales representative, sales person, and engineer having the most significant drops, which suggests that they tend to have a lower sleep quality.

-   Sleep Duration, Physical Activity Level, and Sleep Apnea have a positive estimate, suggesting that longer sleep duration, higher level of physical activity (in minutes), and possessing sleep apnea are associated with a higher quality of sleep. Sleep apnea being associated with higher quality of sleep is contradictory, so there may be improvements that need to be made to the model.

-   Higher stress levels are associated with a decrease in sleep quality.

\quad\quad The multiple R-squared value is 0.9668, which indicates that the model explains about 96.68% of the variance in sleep quality suggesting a good model fit.

### Further Analysis of the Model

```{r}
# Diagnostic plots for regression
par(mfrow = c(1,1))
plot(sleep.model)
```

\quad\quad The diagnostic plots above indicate that the linear assumption might not hold due to the high leverages/outliers, deviations from the diagonal line in the Q-Q Plot, and the obvious trend in the Residuals vs. Fitted Plot. This means that we will have to consider other models because the data set could be too complex to use the linear regression model, or further clean up the model to exclude any potential outliers.

## Tree-Based Models

### Why we Chose this Model

\quad\quad We opted for a decision (regression) tree as our second modeling approach due to its ability to visualize and interpret data effectively, which aligns closely with human decision-making processes compared to linear regression. However, it's important to note that decision trees often have lower predictive accuracy compared to regression models. To address this limitation, we employed pruning techniques to refine the tree and enhance our model's performance.

### Our Formula

\textit{Quality of Sleep} $\sim$ \textit{Gender} + \textit{Age} + \textit{Occupation} + \textit{Sleep Duration} + \textit{Physical Activity Level} + \textit{Stress Level} + \textit{BMI Category} + \textit{Heart Rate} + \textit{Daily Steps} + \textit{Daily Steps} + \textit{Sleep Disorder}

### Regression Tree

# Conclusion

Talk about what we found is the best option, yada yada yada

# References

\text{L. Tharmalingam "Sleep Health and Lifestyle Dataset", Kaggle.} [Link](https://www.kaggle.com/datasets/uom190346a/sleep-health-and-lifestyle-dataset)
