---
title: "Quality Of Sleep Report"
format: pdf
authors: Suchanut Namcharoen, Ashish Job, Chloe Li, Emma Pham, Jaden Thai 
editor: visual
---

## Introduction

As college students, we generally do not get enough sleep. As a result, we thought it'd be interesting to

includeee:

1.  why this dataset ?
2.  what questions are we trying to answer?
    -   Can we predict the quality of sleep (1-10) based on lifestyle factors such as blah blah blah?
3.  which models will we be using ?

## About the Data

The **Sleep Health and Lifestyle Data set** contains 400 rows and 13 columns (400 observations and 13 variables) of data that relates to an individual's lifestyle and sleep habits. The variables are as follows:

-   Person ID: An identifier for each individual.
-   Gender: The gender of the person (Male/Female).
-   Age: The age of the person in years.
-   Occupation: The occupation or profession of the person.
-   Sleep Duration (hours): The number of hours the person sleeps per day.
-   Quality of Sleep (scale: 1-10): A subjective rating of the quality of sleep, ranging from 1 to 10.
-   Physical Activity Level (minutes/day): The number of minutes the person engages in physical activity daily.
-   Stress Level (scale: 1-10): A subjective rating of the stress level experienced by the person, ranging from 1 to 10.
-   BMI Category: The BMI category of the person (e.g., Underweight, Normal, Overweight).
-   Blood Pressure (systolic/diastolic): The blood pressure measurement of the person, indicated as systolic pressure over diastolic pressure.
-   Heart Rate (bpm): The resting heart rate of the person in beats per minute.
-   Daily Steps: The number of steps the person takes per day.
-   Sleep Disorder: The presence or absence of a sleep disorder in the person (None, Insomnia, Sleep Apnea).
    -   None: The individual does not exhibit any specific sleep disorder.

    -   Insomnia: The individual experiences difficulty falling asleep or staying asleep, leading to inadequate or poor-quality sleep.

    -   Sleep Apnea: The individual suffers from pauses in breathing during sleep, resulting in disrupted sleep patterns and potential health risks.

## Data Cleaning

```{r}
# Importing the dataset and viewing summary
data <- read.csv("Sleep_health_and_lifestyle_dataset.csv")
summary(data)
```

To clean up the data set, we evaluated the data set and determined which variables could possibly interfere with the accuracy of the model. We removed PersonID when creating the model, since that represents each observation and would not be used as neither a predictor nor response. The second column we removed was Blood Pressure because we were unable to accurately interpret the blood pressure measurements.

\
We then converted the remaining categorical variables, which included Occupation, Gender, Sleep Disorder, and BMI Category.

```{r}
data$Occupation <- as.factor(data$Occupation) 
data$Gender <- as.factor(data$Gender)
data$Sleep.Disorder <- as.factor(data$Sleep.Disorder)
data$BMI.Category <- as.factor(data$BMI.Category)
```

## Linear Regression Model

We decided to first approach the data set using a multiple linear regression model because we wanted to inference the quality of sleep (1-10), a quantitative response variable, using multiple predictors. Furthermore, linear regression models are generally more straightforward to understand and interpret, so we can quickly identify which predictors are statistically significant and exactly how they affect the response variable.

```{r}
# Creating the multiple linear regression model
sleep.model <- lm(Quality.of.Sleep ~.-Blood.Pressure-Person.ID, data = data)
summary(sleep.model)
```

Based on the summary of the linear regression model, we can observe that multiple predictors are statistically significant (low p-values) as indicated by the asterisks on the right side of the predictors. From these predictors, we notice that:

-   Gender Male has a positive estimate (increasing the quality of sleep by approximately 0.56), indicating that males generally have a better quality of sleep compared to females.

-   Each additional year of age is associated with a a 0.057 increase in quality of sleep, suggesting that sleep quality might improve with age.

-   All of the occupations have a negative estimate, with sales representative, sales person, and engineer having the most significant drops, which suggests that they tend to have a lower sleep quality.

-   Sleep Duration, Physical Activity Level, and Sleep Apnea have a positive estimate, suggesting that longer sleep duration, higher level of physical activity (in minutes), and possessing sleep apnea are associated with a higher quality of sleep. Sleep apnea being associated with higher quality of sleep is contradictory, so there may be improvements that need to be made to the model.

-   Higher stress levels are associated with a decrease in sleep quality.

The multiple R-squared value is 0.9668, which indicates that the model explains about 96.68% of the variance in sleep quality suggesting a good model fit.

### Further Analysis of the Model

```{r}
# Diagnostic plots for regression
par(mfrow = c(1,1))
plot(sleep.model)
```

The diagnostic plots above indicate that the linear assumption might not hold due to the high leverages/outliers, deviations from the diagonal line in the Q-Q Plot, and the obvious trend in the Residuals vs. Fitted Plot. This means that we will have to consider other models because the data set could be too complex to use the linear regression model, or further clean up the model to exclude any potential outliers.

## Random Forest Model
